Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka

### 2025-01-08 analysis/paper.qmd
Revise species list

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 56 plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  Species ranges are predicted to have been on significantly smaller in the 
  Early Holocene compared to present conditions, with many species that are 
  found throughout the region's 'hilly flanks' today indicated to have much more
  restricted distributions centered on the Levant, Cyprus and Western Anatolia. 
  Ranges shrunk by on average c. 10% from the terminal Pleistocene to the Early 
  Holocene, with a further c. 10% reduction for the duration of the Younger 
  Dryas. However, modelled ranges do not reliably predict the observed 
  occurrence of specific species at archaeological sites. [...]
  The regional 
  ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

* [x] Stop modelling
- [ ] First draft:
  - [x] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [x] Results
  - [ ] Discussion
    - [x] General trends
    - [ ] Case studies
    - [x] Comparison with archaeobot
  - [ ] Conclusion
- [ ] Figures
  - [x] fig-region
  - [x] fig-occ
  - [ ] tbl-occ-count (needs pdf cleanup)
  - [ ] tbl-predictors
  - [ ] tbl-model-metrics
- [ ] Outstanding TODOs
- [ ] Add missing references
- [ ] Copyedit
- [ ] Appendix with all hindcast predictions (HTML?)
- [ ] Clean up package code
- [ ] Final proofread

@@

neolithic <- c("PPNA", "EPPNB", "MPPNB", "LPPNB/C", "Pottery Neolithic")

climate_periods <- tibble(
  period = c("Bølling-Allerød", "Younger Dryas", "Early Holocene", "Current"),
  code = c("ba", "yds", "eh", "cur"),
  label = fct_inorder(c("BA (14.7-12.9 ka)", "YDS (12.9-11.7 ka)", 
                      "EH (11.7-8.3 ka)", "current")),
  start_bp = c(14700, 12900, 11700, NA),
  end_bp = c(12900, 11700, 8300, NA)
)
```

```{r define-parameters}
# Minimum number of GBIF occurrences for modelling
min_n_occ <- 50

@@

```{r data-archaeo-flora}
# Flora present at archaeological sites
archaeo_flora <- read_tsv(
  raw_data("swasia_neolithic_flora.tsv"),
  col_types = cols(
    source = col_character(),
    source_id = col_character(),
    source_site_name = col_character(),
    site_name = col_character(),
    latitude = col_double(),
    longitude = col_double(),
    phase = col_character(),
    phase_description = col_character(),
    phase_code = col_character(),
    age_start = col_integer(),
    age_end = col_integer(),
    taxon_source = col_character(),
    n = col_double(),
    prop = col_double(),
    reference = col_character(),
    taxon_detail = col_character(),
    taxon = col_character(),
    genus = col_character(),
    family = col_character(),
    category = col_character(),
    founder_crop = col_character(),
    edibility = col_character(),
    grass_type = col_character(),
    legume_type = col_character()
  )
) |>
  rename(taxon_group = "taxon", taxon = "taxon_detail") |>
  # Assign to archaeological periods
  mutate(
    age_mid = age_end + ((age_start - age_end) / 2),
    period = cut(
      -age_mid, 
      breaks = -c(archaeo_periods$start_bp, 0), 
      labels = archaeo_periods$period,
      ordered_result = TRUE
    ),
    climate_period = cut(
      -age_mid, 
      breaks = -c(climate_periods[climate_periods$code != "cur",]$start_bp, 0), 
      labels = climate_periods[climate_periods$code != "cur",]$code,
      ordered_result = TRUE
    )
  ) ->
  archaeo_flora

n_assemb_neolithic <- archaeo_flora |> 
  distinct(site_name, phase_code, period) |> 
  filter(period %in% neolithic) |> 
  nrow()

n_site_neolithic <- archaeo_flora |> 
  filter(period %in% neolithic) |> 
  distinct(site_name) |> 
  nrow()
```

@@

```{r fig-region}
#| fig-cap: Study region (red) and archaeological sites used to generate modelled flora. Solid circles indicate Neolithic sites.
ggplot() + 
  annotation_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "lightblue") +
  layer_spatial(w_asia, colour = "red", fill = NA) +
  layer_spatial(archaeo_sites, shape = 1) +
  layer_spatial(filter(archaeo_sites, period %in% neolithic), shape = 16) +
  scale_linewidth_identity() +
  coord_sf(crs = w_asia_albers)
```

<!-- TODO define study region and period -->

```{r data-flora}
if (!file_exists(derived_data("flora.tsv"))) {
  message("Regenerating flora...")
  # Subset of flora to be modelled
  archaeo_flora |>
    filter(period %in% neolithic) |>
    group_by(taxon) |>
    summarise(
      n_present = length(unique(phase_code)),
      p_present = n_present / n_assemb_neolithic,
      .groups = "drop"
    ) |>
    filter(n_present > 2) |>
    # Exclude taxa not identified to species
    filter(!str_ends(taxon, "sp.")) ->
    flora

@@

We began by considering `r nrow(flora)` distinct taxa (@tbl-occ-count) - all the identifiable species known to be present at more than three Neolithic (c. 11.7–6.5 ka) sites in West Asia, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
<!-- TODO: something about not distinguishing between source of plant remains? -->
Taxonomic names were resolved to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-occ-count).
Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where different.

@@

```{r data-occ-count}
if (!file_exists(derived_data("flora_occ_count.tsv"))) {
  message("Reacquiring GBIF occurrence counts...")
  flora <- flora |>
    mutate(
      n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                            hasCoordinate = TRUE, 
                                            hasGeospatialIssue = FALSE,
                                            geometry = bbox_wkt(w_asia)))
    )
  write_tsv(flora, derived_data("flora_occ_count.tsv"))
}
flora <- read_tsv(
  derived_data("flora_occ_count.tsv"),
  col_types = cols(
    taxon = col_character(),
    taxon_detail = col_character(),
    n_present = col_double(),
    p_present = col_double(),
    n_occ = col_double()
  )
)
```

```{r tbl-occ-count}
#| tbl-cap: "Recorded occurrence of flora considered in this study in Neolithic and contemporary West Asia"
flora |>
  mutate(
    exclude = (n_occ < min_n_occ) | (taxon == "Avena sterilis"),
    label = if_else(
      !is.na(taxon_detail),
      glue::glue("*{taxon}*\n\n<small>incl. {taxon_detail}</small>"),
      paste0("*", taxon, "*")
    )
  ) |>
  select(label, n_present, p_present, n_occ, exclude) |>
  arrange(-n_present) |>
  gt() |>
  cols_label(
    label = "Taxon",
    n_present = "Neolithic",
    p_present = "(%)",
    n_occ = "Present"
  ) |>
  tab_spanner("Occurrences", columns = c(n_present, p_present, n_occ)) |>
  # cols_merge_n_pct(n_present, p_present) |>
  cols_hide(exclude) |>
  tab_footnote(
    "Excluded from modelling due to sample size", 
    cells_body(columns = c(n_occ), rows = exclude),
    placement = "right"
  ) |>
  cols_align("left", c(label)) |>
  cols_align("right", c(n_present, p_present, n_occ)) |>
  cols_width(label ~ pct(50), n_present ~ 10, p_present ~ 10, n_occ ~ 20) |>
  fmt_percent(p_present, decimals = 0) |>
  fmt_markdown(label) |>
  sub_missing() |>
  tab_options(
    footnotes.marks = "*",
    table.font.size = 11,
    table.width = pct(100),
    latex.use_longtable = TRUE
  )
```

Georeferenced occurrence data was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
GBIF occurrences marked has having imprecise or duplicate coordinates were excluded from the training dataset, as were fossil records.
Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded `r english(nrow(filter(flora, n_occ > min_n_occ)))` taxa with less than `r min_n_occ` occurrences in West Asia, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
We also excluded one taxon (*Avena sterilis*) with over 50,000 occurrences, as this would have been computationally prohibitive and we were uncertain what account for such a disproportionately high number of records.

```{r data-filter-flora}
flora <- filter(flora, n_occ > min_n_occ)
# Why are there 50k occurrences of Avena sterilis??
flora <- filter(flora, taxon != "Avena sterilis")
```

```{r data-occ}
# TODO: further tidying. Ideas:
# - https://data-blog.gbif.org/post/gbif-filtering-guide/
# - https://cran.r-project.org/web/packages/CoordinateCleaner/vignettes/Cleaning_GBIF_data_with_CoordinateCleaner.html
# - https://www.r-bloggers.com/2021/03/downloading-and-cleaning-gbif-data-with-r/
if (!file_exists(derived_data("flora_occ.tsv"))) {
  message("Reacquiring GBIF occurrences...")
  flora <- mutate(
    flora,
    occ = map(
      taxon, gbif_data, 
      hasCoordinate = TRUE, hasGeospatialIssue = FALSE, geometry = w_asia, 
      limit = 100000, .progress = TRUE
    ),
    occ = map(occ, gbif_drop_imprecise_coords, threshold = 5000),
    occ = map(occ, gbif_drop_fossils),
    occ = map(occ, gbif_drop_duplicate_coords),
    occ = map(occ, gbif_standardise_names),
    occ = map(occ, \(x) {
      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
             latitude)
    })
  )
  write_tsv(unnest(flora, cols = c(occ)), derived_data("flora_occ.tsv"))
}
flora <- read_tsv(
  derived_data("flora_occ.tsv"),
  col_types = cols(
    taxon = col_character(),
    taxon_detail = col_character(),
    n_present = col_double(),
    p_present = col_double(),
    n_occ = col_double(),
    gbif_key = col_double(),
    scientific_name = col_character(),
    genus = col_character(),
    species = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
) |>
  nest(occ = gbif_key:latitude)
```

```{r fig-occ-map, fig.cap=paste0("Georeferenced occurrence records from West Asia for the ", nrow(flora), " modelled taxa (N=", nrow(unnest(flora, occ)))}
flora |>
 unnest(occ) |>
 distinct(latitude, longitude) |>
 st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
 ggplot() +
 annotation_spatial(ne_land, fill = "#ffffff") +
 geom_sf(shape = ".") +
 layer_spatial(w_asia, fill = NA) +
 coord_sf(crs = w_asia_albers, default_crs = latlong)
```

@@

<!-- TODO: rewrite -->
Occurrence data only tells us where a species is present;
there is rarely definitive information on where the species is *not* found.
We therefore need to generate random background points or "pseudo-absences" to feed to the model.
There are several ways to do this. 
We follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (:10000) random sample of points, weighted equally against the presences in the regression.
@ValaviEtAl2022 also recommend using a very large background sample for random forest models.

@@

@@

```{r niche-model}
# Random Forest Niche Model
# With hyperparameters adapted from the example of a down-sampled RF model given 
# by https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
niche_model <- rand_forest(mode = "classification", trees = 1000) |>
  set_engine(
    "ranger", 
    sample.fraction = c(sum(as.logical(.y())) / n_background, 1),
    # probability = TRUE
  )

@@

```{r data-model-assessment}
flora <- mutate(
  flora,
  pred = map(niche, collect_predictions),
  niche_test_metrics = map(niche, collect_metrics),
  sensitivity = map(pred, sens, truth = present, estimate = .pred_class,
                    event_level = "second"), # reference present=TRUE
  niche_test_metrics = map2(niche_test_metrics, sensitivity, bind_rows)
) |>
  select(-sensitivity)

niche_test_metrics <- flora |>
  unnest(niche_test_metrics) |>
  summarise(
    min = min(.estimate),
    max = max(.estimate),
    mean = mean(.estimate),
    sd = sd(.estimate),
    .by = c(.metric)
  )
```

@@

We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition.
Model accuracy (proportion of correctly classified presence and background samples) ranged between `r percent(filter(niche_test_metrics, .metric == "accuracy")$min)` and `r percent(filter(niche_test_metrics, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_metrics, .metric == "accuracy")$mean)`.
Sensitivity (proportion of correctly classified presence samples) ranged between `r percent(filter(niche_test_metrics, .metric == "sens")$min)` and `r percent(filter(niche_test_metrics, .metric == "sens")$max)`, with an average of `r percent(filter(niche_test_metrics, .metric == "sens")$mean)`.
The area under the models' receiver operating characteristic curves (ROC-AUC) was in the range of `r round(filter(niche_test_metrics, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_metrics, .metric == "accuracy")$sd, 3)`
Model sensitivity is loosely correlated with the number of occurrences available for training (@fig-model-sensitivity-vs-n_occ), with the worst-performing models all having less than 300 recorded occurrences: `r knitr::combine_words(paste0("*", filter(unnest(flora, niche_test_metrics), .metric == "sens" & .estimate < 0.85)$taxon, "*"))`.
Test metrics and ROC curves for the individual models are included in the appendix.

```{r niche-archaeo}
# TODO: factor out
# TODO: what about negatives?
archaeo_flora_predictions <-

archaeo_flora |>
  filter(taxon %in% flora$taxon) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = latlong) |>
  mutate(
    prediction = map(taxon, \(x) filter(flora, taxon == x)$palaeodist[[1]]),
    prediction = map2(prediction, geometry, function(pred, geom) {
      sf_point <- st_transform(st_sf(st_sfc(geom, crs = latlong)), w_asia_albers)
      st_extract(pred, sf_point)
    }),
    prediction = map(prediction, as_tibble),
    prediction = map2(prediction, climate_period, function(pred, per) {
      filter(pred, as.character(period) == per)
    }),
    prediction = map(prediction, select,
                     pred_present = present, pred_absent = absent)
  ) |>
  unnest(c(prediction))

@@

# Results

@@

```{r fig-range-density}
#| fig-cap: Distribution of predicted species ranges by period. Dashed lines indicate the median range.
flora_pred_area_present_summary <- flora |>
  unnest(area_present) |>
  summarise(med_area_present = median(area_present) / 1e6, .by = period) |>
  mutate(
    period_label = map(period, \(x) filter(climate_periods, code == x)$label),
    period_label = unlist(period_label),
  )

>>>>

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 81<!-- TODO update --> plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  The models predict significant reductions and/or shifts in species ranges in
  the terminal Pleistocene and Early Holocene compared to present conditions. 
  Many species that are found throughout the region's 'hilly flanks' today 
  are indicated to have much more restricted distributions centered on the 
  Levant, Cyprus and Western Anatolia. In addition, overall ranges shrunk by an 
  average of c. 25% from the terminal Pleistocene to the Early Holocene. 
  Modelled ranges do not reliably predict the observed  occurrence of specific 
  species at archaeological sites,
  however [...]
  The regional ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

* [x] Stop modelling
- [ ] First draft:
  - [x] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [x] Results
  - [ ] Discussion
    - [x] General trends
    - [ ] Case studies
    - [x] Comparison with archaeobot
  - [ ] Conclusion
- [ ] Figures
  - [x] fig-region
  - [x] fig-occ
  - [ ] tbl-occ-count (needs pdf cleanup)
  - [ ] tbl-predictors
  - [ ] tbl-model-metrics
- [ ] Outstanding TODOs
- [ ] Add missing references
- [ ] Copyedit
- [x] Appendix with all hindcast predictions (HTML?)
- [ ] Clean up package code
- [ ] Final proofread

@@

epipal <- c("Late Epipal.")
ppn <- c("PPNA", "EPPNB", "MPPNB", "LPPNB/C")

climate_periods <- tibble(
  period = c("Bølling-Allerød", "Younger Dryas", "Early Holocene", "Current"),
  code = c("ba", "yds", "eh", "cur"),
  date_range = c("(14.7-12.9 ka)", "(12.9-11.7 ka)", "(11.7-8.3 ka)", ""),
  label = fct_inorder(paste(toupper(code), date_range)),
  long_label = fct_inorder(paste(period, date_range)),
  start_bp = c(14700, 12900, 11700, NA),
  end_bp = c(12900, 11700, 8300, NA)
)
```

```{r define-parameters}
# Minimum number archaeological occurrences for modelling
min_n_archaeo <- 1

# Minimum number of present occurrences for modelling
min_n_occ <- 50

@@

```{r data-archaeo-flora}
# Flora present at archaeological sites
archaeo_flora <- read_tsv(
  raw_data("swasia_neolithic_flora.tsv"),
  col_types = cols(
    source = col_character(),
    source_id = col_character(),
    source_site_name = col_character(),
    site_name = col_character(),
    latitude = col_double(),
    longitude = col_double(),
    phase = col_character(),
    phase_description = col_character(),
    phase_code = col_character(),
    age_start = col_integer(),
    age_end = col_integer(),
    taxon_source = col_character(),
    n = col_double(),
    prop = col_double(),
    reference = col_character(),
    taxon_detail = col_character(),
    taxon = col_character(),
    genus = col_character(),
    family = col_character(),
    category = col_character(),
    founder_crop = col_character(),
    edibility = col_character(),
    grass_type = col_character(),
    legume_type = col_character()
  )
) |>
  rename(taxon_group = "taxon", taxon = "taxon_detail") |>
  # Assign to archaeological periods
  mutate(
    age_mid = age_end + ((age_start - age_end) / 2),
    period = cut(
      -age_mid, 
      breaks = -c(archaeo_periods$start_bp, 0), 
      labels = archaeo_periods$period,
      ordered_result = TRUE
    ),
    climate_period = cut(
      -age_mid, 
      breaks = -c(climate_periods[climate_periods$code != "cur",]$start_bp, 0), 
      labels = climate_periods[climate_periods$code != "cur",]$code,
      ordered_result = TRUE
    )
  ) |>
  # Filter to Epipal/Neo
  filter(period %in% epipal | period %in% ppn)

n_archaeo_assemb <- nrow(distinct(archaeo_flora, site_name, phase_code))
n_archaeo_site <- nrow(distinct(archaeo_flora, site_name))
```

@@

```{r fig-region}
#| fig-cap: Late Epipalaeolithic & Pre-Pottery Neolithic archaeological sites used to generate modelled flora
ggplot() + 
  annotation_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "lightblue") +
  layer_spatial(w_asia, fill = NA) +
  layer_spatial(archaeo_sites, size = 0.5) +
  scale_linewidth_identity() +
  coord_sf(crs = w_asia_albers)
```

```{r data-flora}
# TODO: nest archaeo flora in flora to make it easier to keep track of taxon transformations
if (!file_exists(derived_data("flora.tsv"))) {
  message("Regenerating flora...")
  # Subset of flora to be modelled
  archaeo_flora |>
    group_by(taxon) |>
    summarise(
      n_present = length(unique(phase_code)),
      p_present = n_present / n_archaeo_assemb,
      .groups = "drop"
    ) |>
    filter(n_present > min_n_archaeo) |>
    # Exclude taxa not identified to species
    filter(!str_ends(taxon, "sp.")) ->
    flora

@@

The aim of our study was to model the biogeography of species relevant to human subsistence economies in West Asia (excluding the Southern Arabian peninsula, see @fig-region) during the archaeological Late Epipalaeolithic (15–11.7 ka) and Pre-Pottery Neolithic (11.7–8.3 ka) periods.
Our starting point was a list of `r nrow(flora)` taxa (@tbl-taxa) comprising the identifiable species observed at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
This was based on dataset of `r n_archaeo_assemb` assemblages collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@COMPAG].
We did not attempt to distinguish between the source of the remains;
archaeobotanical assemblages are subject to a variety of preservational and recovery biases, so by no means were all the species on our list consumed or even deliberately collected by people.
However, we assume that there presence at a site of human settlement at least implies that they were part of the wider ecosystem that supported habitation there.

Taxonomic names were resolved to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-taxa).
Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where different.

@@

```{r data-n-occ} 
if (!file_exists(derived_data("flora_n_occ.tsv"))) {
  message("Reacquiring GBIF occurrence counts...")
  flora <- flora |>
    mutate(
      n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                            hasCoordinate = TRUE, 
                                            hasGeospatialIssue = FALSE,
                                            geometry = bbox_wkt(w_asia)))
    )
  write_tsv(flora, derived_data("flora_n_occ.tsv"))
}

flora <- read_tsv(
  derived_data("flora_n_occ.tsv"),
  col_types = cols(
    taxon = col_character(),
    taxon_detail = col_character(),
    n_present = col_double(),
    p_present = col_double(),
    n_occ = col_integer()
  )
)

# Filter small samples
# ...and one very large one (why are there 50k occurrences of Avena sterilis?)
flora_all <- flora
flora <- flora |>
  filter(n_occ > min_n_occ) |>
  filter(taxon != "Avena sterilis")
```

```{r data-occ}
# Cleaning pipeline based on:
# - https://data-blog.gbif.org/post/gbif-filtering-guide/
if (!file_exists(derived_data("flora_occ.tsv"))) {
  message("Reacquiring GBIF occurrences...")
  flora <- mutate(
    flora,
    occ = map(
      taxon, gbif_data, 
      hasCoordinate = TRUE, hasGeospatialIssue = FALSE, geometry = w_asia, 
      limit = 100000, .progress = TRUE
    ),
    occ = map(occ, filter, occurrenceStatus  == "PRESENT"),
    occ = map(occ, gbif_drop_imprecise_coords, threshold = 1000),
    occ = map(occ, gbif_drop_fossils),
    occ = map(occ, gbif_drop_duplicate_coords),
    occ = map(occ, gbif_standardise_names),
    occ = map(occ, \(x) {
      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
             latitude)
    }),
    # Update occ count after cleaning
    n_occ = map_int(occ, nrow)
  )

  # Refilter based on cleaned records
  flora <- filter(flora, n_occ > min_n_occ)

  write_tsv(unnest(flora, cols = c(occ)), derived_data("flora_occ.tsv"))
}

flora <- read_tsv(
  derived_data("flora_occ.tsv"),
  col_types = cols(
    taxon = col_character(),
    taxon_detail = col_character(),
    n_present = col_double(),
    p_present = col_double(),
    n_occ = col_integer(),
    gbif_key = col_double(),
    scientific_name = col_character(),
    genus = col_character(),
    species = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
) |>
  nest(occ = gbif_key:latitude)
```

Georeferenced occurrence data for West Asia was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
GBIF was cleaned to removed fossil occurrences, recorded absences, and records with missing, imprecise (>1 km uncertainty), or invalid coordinates.
Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded `r english(nrow(filter(flora, n_occ > min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
We also excluded one taxon (*Avena sterilis*) with over 50,000 occurrences, as this would have been computationally prohibitive and we were uncertain what accounted for such a disproportionately high number of records.

```{r fig-occ-map, fig.cap=paste0("Georeferenced occurrence records from West Asia used to train models (N=", nrow(unnest(flora, occ)))}
flora |>
 unnest(occ) |>
 distinct(latitude, longitude) |>
 st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
 ggplot() +
 annotation_spatial(ne_land, fill = "#ffffff") +
 geom_sf(shape = ".") +
 layer_spatial(w_asia, fill = NA) +
 coord_sf(crs = w_asia_albers, default_crs = latlong)
```

@@

<!-- TODO: rewrite -->
Occurrence data only tells us where a species is present;
there is rarely definitive information on where the species is *not* found.
We therefore need to generate random background points or "pseudo-absences" to feed to the model.
There are several ways to do this. 
We follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (≈10000) random sample of points, weighted equally against the presences in the regression.
@ValaviEtAl2022 also recommend using a large background sample for random forest models.

@@

```{r tbl-predictors}
# TODO
```

@@

```{r niche-model}
# Random Forest Niche Model
# With hyperparameters adapted from the example of a down-sampled RF model given 
# by https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
niche_model <- rand_forest(mode = "classification", trees = 1000) |>
  set_engine(
    "ranger", 
    sample.fraction = c(sum(as.logical(.y())) / n_background, 1),
    importance = "impurity"
  )

@@

```{r data-model-assessment}
flora <- mutate(
  flora,
  niche_test_metrics = map(niche, collect_metrics),
  sensitivity = map(pred, sens, truth = present, estimate = .pred_class,
                    event_level = "second"), # reference present=TRUE
  niche_test_metrics = map2(niche_test_metrics, sensitivity, bind_rows)
) |>
  select(-sensitivity)

niche_test_summary <- flora |>
  unnest(niche_test_metrics) |>
  summarise(
    min = min(.estimate),
    max = max(.estimate),
    mean = mean(.estimate),
    sd = sd(.estimate),
    .by = c(.metric)
  )
```

@@

We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition.
Model accuracy (proportion of correctly classified presence and background samples) ranged between `r percent(filter(niche_test_summary, .metric == "accuracy")$min)` and `r percent(filter(niche_test_summary, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_summary, .metric == "accuracy")$mean)`.
Sensitivity (proportion of correctly classified presence samples) ranged between `r percent(filter(niche_test_summary, .metric == "sens")$min)` and `r percent(filter(niche_test_summary, .metric == "sens")$max)`, with an average of `r percent(filter(niche_test_summary, .metric == "sens")$mean)`.
The area under the models' receiver operating characteristic curves (ROC-AUC) was in the range of `r round(filter(niche_test_summary, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_summary, .metric == "accuracy")$sd, 3)`
Model sensitivity is loosely correlated with the number of occurrences available for training (@fig-model-sensitivity-vs-n_occ), with the worst-performing models all having less than 300 recorded occurrences: `r knitr::combine_words(paste0("*", filter(unnest(flora, niche_test_metrics), .metric == "sens" & .estimate < 0.85)$taxon, "*"))`.
Test metrics and ROC curves for the individual models are included in the appendix.

```{r niche-archaeo}
# TODO: need to deal with taxon variants / GBIF normalisation
# TODO: factor out
# TODO: what about negatives?
archaeo_flora_predictions <-  archaeo_flora |>
  filter(taxon %in% flora$taxon) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = latlong, remove = FALSE) |>
  mutate(
    prediction = map(taxon, \(x) filter(flora, taxon == x)$palaeodist[[1]]),
    prediction = map2(prediction, geometry, function(pred, geom) {
      sf_point <- st_transform(st_sf(st_sfc(geom, crs = latlong)), w_asia_albers)
      st_extract(pred, sf_point)
    }),
    prediction = map(prediction, as_tibble),
    prediction = map2(prediction, climate_period, function(pred, per) {
      filter(pred, as.character(period) == per)
    }),
    prediction = map(prediction, select,
                     pred_present = present, pred_absent = absent)
  ) |>
  unnest(c(prediction)) |>
  as_tibble() |>
  select(-geometry)

write_tsv(archaeo_flora_predictions, derived_data("archaeo_flora_predictions.tsv"))

@@

```{r tbl-taxa}
#| tbl-cap: "Summary of species modelled"
# TODO missing excluded species
# TODO: redo with:
# - alt. taxons as footnotes
# - check boxes of which (climate?) periods observed in ?
flora |>
  # Summarise archaeo occurrences by period

  # Extract model metrics
  unnest(c(niche_test_metrics)) |>
  select(-.estimator, -.config) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  # Excluded samples
  mutate(exclude = (n_occ < min_n_occ) | (taxon == "Avena sterilis")) |>
  # Prepare table
  select(taxon, n_present, n_occ, exclude, accuracy, roc_auc, sens) |>
  arrange(taxon) |>
  # Table
  gt() |>
  cols_label_with(fn = str_to_sentence) |>
  cols_label(
    n_present = "Archaeological",
    n_occ = "Present (GBIF)",
    roc_auc = "ROC-AUC",
    sens = "Sensitivity"
  ) |>
  tab_spanner("Occurrences", columns = c(n_present, n_occ)) |>
  tab_spanner("Model", columns = accuracy:sens) |>
  cols_hide(exclude) |>
  tab_footnote(
    "Excluded from modelling due to sample size", 
    cells_body(columns = c(n_occ), rows = exclude),
    placement = "right"
  ) |>
  cols_align("right") |>
  cols_align("left", c(taxon)) |>
  fmt_number(accuracy:sens) |>
  sub_missing() |>
  tab_style(cell_text(style = "italic"), cells_body(taxon)) |>
  tab_options(
    footnotes.marks = "*",
    table.font.size = 11,
    table.width = pct(100),
    latex.use_longtable = TRUE
  )
```

@@

```{r data-pred-range-summary}
flora_pred_area_present_summary <- flora |>
  unnest(area_present) |>
  summarise(med_area_present = median(area_present) / 1e6, .by = period) |>
  mutate(
    period_label = map(period, \(x) filter(climate_periods, code == x)$label),
    period_label = unlist(period_label),
  )

n_reduced_range <- flora |>
  unnest(area_present) |>
  pivot_wider(names_from = period, values_from = area_present) |>
  filter(eh < ba) |>
  nrow()

n_reduced_range_10 <- flora |>
  unnest(area_present) |>
  pivot_wider(names_from = period, values_from = area_present) |>
  filter(eh <= ba * 0.9) |>
  nrow()
```

```{r fig-pred-range-density}
#| fig-cap: Distribution of predicted species ranges by period. Dashed lines indicate the median range.
flora |>
  unnest(area_present) |> 
  mutate(
    period_label = map(period, \(x) filter(climate_periods, code == x)$label),
    period_label = unlist(period_label),
    area_present = area_present / 1e6 # m2 to km2
  ) |> 
  ggplot() + 
  facet_wrap(vars(period_label), ncol = 1, strip.position = "left") + 
  geom_density(aes(area_present)) +
  geom_vline(aes(xintercept = med_area_present), flora_pred_area_present_summary) +
  scale_x_log10(name = "Predicted range km²", labels = label_number_auto()) + 
  scale_y_continuous(name = NULL, labels = NULL) +
  theme_minimal_vgrid() +
  theme(
    strip.text.y.left = element_text(angle = 0, hjust = 0),
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```
